---
title: "DS5500_Homework1"
author: "Noah DeMoes"
date: "October 4, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
```

## Problem 1

I first scored a 31%, but I took it again and received a 100%.

The question that surprised me the most was about the expected number of children in 2100.
The answer is 2 billion which is the same number of children that exist today. I thought that this prediction is interesting given life expectancy is expected to increase. Perhaps, the model is taking into account the Earth's carry capacity. As the number of adults increases, there is prediction that either people will stop having as many children.

The question I selected was problem 2.

In the last 20 years the proportion of people living in extreme poverty worldwide, has...?

```{r}
# Read in files
gapfiles<-list.files()

# Get all csv files
alllfiles<-list.files(gapfiles[1],pattern = '*.csv')

#Get only the disaster files that relate to deaths
Disasterfiles<-c()
i=1
for (f in alllfiles){
  a=grepl('extreme_poverty',f)

  if( a == TRUE){
    Disasterfiles[i]=f
    i=i+1
    }

}


###Read in extreme poberty file
for (f in Disasterfiles){
  df=read.csv(paste0(gapfiles[1],'/',Disasterfiles[1]))
}

df[is.na(df)]<-remove()
dflast20<-df%>%subset(df$time>=1998)
#### Summarize

poverty_summary<-dflast20 %>% group_by(time) %>% summarise(poverty=mean(extreme_poverty_percent_people_below_190_a_day))#### Visualization


ggplot(data=poverty_summary)+geom_line(aes(x=poverty_summary$time,y=poverty_summary$poverty))+geom_point(aes(x=poverty_summary$time,y=poverty_summary$poverty))+
ggtitle('Percentage of World in Extreme Poverty by Year since 1998')+xlab('Year')+ylab("Percentage of Population in Extreme Poverty")
difference=poverty_summary$poverty[1]-poverty_summary$poverty[20]
print(difference)
```

The average percentage of people in the World that live in extreme poverty has decreased by 20.80% from 199 to 2017.


## Problem 2


```{r }
library(countrycode)
GDPfiles<-c()
i=1
for (f in alllfiles){
  a=grepl('income_per_person_gdppercapita_ppp_inflation_adjusted',f)
  b=grepl('us',f)

  if( a == TRUE){
    GDPfiles[i]=f
    i=i+1
    }

}

for (f in GDPfiles){
  df=read.csv(paste0(gapfiles[1],'/',GDPfiles[1]))
}

df$geo=as.character(lapply(df$geo,toupper))
df$continent<-countrycode(sourcevar = as.character(df$geo),origin = 'iso3c',destination = 'continent')





gdpsummary<-df%>%group_by(continent,time)%>%summarise(gdp=mean(income_per_person_gdppercapita_ppp_inflation_adjusted))



ggplot(data=gdpsummary)+geom_line(aes(x=time,y=gdp,group=continent,colour=continent))+
  ggtitle('GDP/capita vs time by Continent')+xlab('Year')+ylab('GDP/Capita')

for (cont in unique(df$continent)){
dg=df[df$continent==cont,]
print(ggplot(data=dg)+geom_line(aes(x=time,y=income_per_person_gdppercapita_ppp_inflation_adjusted,group=geo,colour=geo))+ggtitle(paste0("GDP/Capita vs year by Countries in ",cont))+xlab("Year")+ylab('GDP/Capita'))}
  
GDPdf=df
```
I first looked at the GDP/Capita for each Continent as a function of time from 1800 to 2050. It appears as though for the first 150 years each continent was on the same tradjectory with only a small variance between continents. After 1975, however, each continent's GDP/capita seperates, with Europe leading, followed by Asia, Americas, Oceania, and Africa. Perhaps the baby boom contributed to the seperation of GDP.

Additionally, I provided graphics of the GDP/Capita by country seperated by continent. Asia, Oceania, and Africa all have one or two countries that spike in GDP/Capita during 1975 while the countries in Europe and Americas seem to follow more stable trends. Perhaps, this means that countries in Europe and the Americas are more dependent on each other and work together more than countries on other continents. However, USA seems to do well despite variations in other countries.

## Problem 3

```{r}
lifemortalfiles<-c()
i=1
for (f in alllfiles){
  a=grepl('child_mortality',f)
  b=grepl('us',f)

  if( a == TRUE){
    lifemortalfiles[i]=f
    i=i+1
    }

}

for (f in lifemortalfiles){
  child_df=read.csv(paste0(gapfiles[1],'/',lifemortalfiles[1]))
}


child_df$geo=as.character(lapply(child_df$geo,toupper))


lifemortalfiles<-c()
i=1

for (f in alllfiles){
  a=grepl('life_expectancy_years',f)
  b=grepl('us',f)

  if( a == TRUE){
    lifemortalfiles[i]=f
    i=i+1
    }

}

for (f in lifemortalfiles){
  lifeexpectancy_df=read.csv(paste0(gapfiles[1],'/',lifemortalfiles[1]))
}
lifeexpectancy_df$geo=as.character(lapply(lifeexpectancy_df$geo,toupper))


p3df=merge(GDPdf,child_df,by=c('geo','time'),all=T)

p3df=remove_missing(p3df)

p3df=merge(p3df,lifeexpectancy_df,by=c('geo','time'),all=T)
p3df=remove_missing(p3df)


gdpsummary<-p3df%>%group_by(continent,time)%>%summarise(gdp=mean(income_per_person_gdppercapita_ppp_inflation_adjusted),life_exp=mean(life_expectancy_years),child_mortality=mean(child_mortality_0_5_year_olds_dying_per_1000_born))



ggplot(data=gdpsummary)+geom_line(aes(x=time,y=life_exp,group=continent,colour=continent))+
  ggtitle('Life Expectancy vs Time by Continent')+xlab('Year')+ylab('Life Expectancy')
ggplot(data=gdpsummary)+geom_line(aes(x=time,y=child_mortality,group=continent,colour=continent))+
  ggtitle('Child Mortality vs Time by Continent')+xlab('Year')+ylab('Child Mortality')

ggplot(data=gdpsummary)+geom_line(aes(x=time,y=gdp,group=continent,colour=continent))+
  ggtitle('GDP/capita vs time by Continent')+xlab('Year')+ylab('GDP/Capita')

```

The main outlier for life expectance occurs during 1914-1919 and 1939-1944, during world war world war I and II. All countries are effected. Besides that period, life expectance has trended higher as time goes increased from 1800 to 2017. There does appear to be a sharp increase in Life Expectancy after 1950 that trends higher. This can be attributed to advancements in healthcare and modern medicine. This conclusion is supported by the trends in child mortality. Child mortality rates year over year changed moderately between 1800-1900. It wasn't until 1900 did they decrease at a higher rate. It should be noted that modern medicine did not arrive at each continent at the same time which attributes to the slower decrease in mortaility rates for AFrica and Asia compared to Europe, Oceania, and the Americas.

Additionally, rate of change for GDP/captia follows relatively the same reciprocal of child mortality rates. It is more robust to life expectance as shown by the lack of change in GDP/capita during the world wars.


## Problem 4


```{r}
cancer<-c()
i=1
for (f in alllfiles){
  a=grepl('malnutr',f)
  b=grepl('us',f)

  if( a == TRUE){
    cancer[i]=f
    i=i+1
    }

}

for (f in cancer){
  malnutr=read.csv(paste0(gapfiles[1],'/',cancer[1]))
}
malnutr$geo=as.character(lapply(malnutr$geo,toupper))



sugar<-c()
i=1
for (f in alllfiles){
  a=grepl('sugar',f)
  b=grepl('us',f)

  if( a == TRUE){
    sugar[i]=f
    i=i+1
    }

}

for (f in sugar){
  sugar_df=read.csv(paste0(gapfiles[1],'/',sugar[1]))
}

sugar_df$geo=as.character(lapply(sugar_df$geo,toupper))

p4df=merge(malnutr,sugar_df,by=c('geo','time'),all=T)
p4df<-remove_missing(p4df)
p4df$continent<-countrycode(sourcevar = as.character(p4df$geo),origin = 'iso3c',destination = 'continent')

p4summary<-p4df%>%group_by(continent,time)%>%summarise(sugar=mean(sugar_per_person_g_per_day),malnut=mean(malnutrition_weight_for_age_percent_of_children_under_5))


ggplot(data=p4summary)+geom_line(aes(x=time,y=malnut,group=continent,colour=continent))+
  ggtitle('Percentage underweight vs time by Continent')+xlab('Year')+ylab('Percent underweight')


ggplot(data=p4summary)+geom_line(aes(x=time,y=sugar,group=continent,colour=continent))+
  ggtitle('amount of sugar vs time by Continent')+xlab('Year')+ylab('amount of sugar')


```
In problem 4 I looked at the relationship between the amount of sugar consumed by continent and the level of malnurishment in the continent. The result of the amount of sugar by continent is hard to distinguish. Every continent experiences an oscillating pattern for the amount of sugar consumed year over year. Because of the oscillation it is hard to draw any meaningful relationship between the amount of sugar and thelevel of malnutrition.

## Problem 5

I utilized static plots.

The advantage of utilizing a static plot is you can select the percise information you would like to present for a given topic. This ability limits the likelihood that an audience will get lost or overwhelmed by too much information. The disadvantage is that it can be hard to convey changes with time for mutliple variables. Additionally, it limits your ability to adapt to a variety of questions, which is one of the advantages of having dynamic plots. Dynamic visualizations enable a research to capture a much wider range of information. Additonally, you can capture changes in two or more variables as a function of another and still maintain understanding.